---
title: "The State of Women’s Paychecks:"
subtitle: "How women’s salaries have compared to men’s across all fifty U.S. states over the past decade"
author: "Emma Oliver"
format: html
editor: visual
---

```{r Load packages and data}
#| include: false
#| 
# Loading Packages
library(ggplot2)
library(rsconnect)
library(ggridges)
library(tidyverse)
library(rnaturalearth)
library(sf)
library(leaflet)
library(dplyr)
library(readr)
library(lubridate)
library(gganimate)
library(gifski)
library(png)

#Loading Data
Iowa <- read_csv("file (3).csv")

Iowa <- Iowa |>
  select(Year, Value)

Kansas <- read_csv("file (4).csv")

Kansas <- Kansas |>
  select(Year, Value)

Kentucky <- read_csv("file (1).csv")

Kentucky <- Kentucky |>
  select(Year, Value)

Louisiana <- read_csv("file (2).csv")

Louisiana <- Louisiana |>
  select(Year, Value)

Indiana <- read_csv("file (5).csv")

Indiana <- Indiana |>
  select(Year, Value)

Wyoming <- read_csv("file (6).csv")

Wyoming <- Wyoming |>
  select(Year, Value)

Maryland <- read_csv("file (7).csv")

Maryland <- Maryland |>
  select(Year, Value)

Massachusetts <- read_csv("file (8).csv")

Massachusetts <- Massachusetts |>
  select(Year, Value)

Mississippi <- read_csv("file (9).csv")

Mississippi <- Mississippi |>
  select(Year, Value)

Missouri <- read_csv("file (10).csv")

Missouri <- Missouri |>
  select(Year, Value)

Montana <- read_csv("file (11).csv")

Montana <- Montana |>
  select(Year, Value)

Utah <- read_csv("file (12).csv")

Utah <- Utah |>
  select(Year, Value)

Vermont <- read_csv("file (13).csv")

Vermont <- Vermont |>
  select(Year, Value)

Virginia <- read_csv("file (14).csv")

Virginia <- Virginia |>
  select(Year, Value)

Washington <- read_csv("file (15).csv")

Washington <- Washington |>
  select(Year, Value)

WestVirginia <- read_csv("file (16).csv")

WestVirginia <- WestVirginia |>
  select(Year, Value)

Michigan <- read_csv("file (17).csv")

Michigan <- Michigan |>
  select(Year, Value)

Minnesota <- read_csv("file (18).csv")

Minnesota <- Minnesota |>
  select(Year, Value)

Maine <- read_csv("file (19).csv")

Maine <- Maine |>
  select(Year, Value)

Wisconsin <- read_csv("file (20).csv")

Wisconsin <- Wisconsin |>
  select(Year, Value)

NewHampshire <- read_csv("file (21).csv")

NewHampshire <- NewHampshire |>
  select(Year, Value)

NewJersey <- read_csv("file (22).csv")

NewJersey <- NewJersey |>
  select(Year, Value)

NewMexico <- read_csv("file (23).csv")

NewMexico <- NewMexico |>
  select(Year, Value)

NorthDakota <- read_csv("file (24).csv")

NorthDakota <- NorthDakota |>
  select(Year, Value)

Ohio <- read_csv("file (25).csv")

Ohio <- Ohio |>
  select(Year, Value)

Oklahoma <- read_csv("file (26).csv")

Oklahoma <- Oklahoma |>
  select(Year, Value)

RhodeIsland <- read_csv("file (27).csv")

RhodeIsland <- RhodeIsland |>
  select(Year, Value)

SouthCarolina <- read_csv("file (28).csv")

SouthCarolina <- SouthCarolina |>
  select(Year, Value)

SouthDakota <- read_csv("file (29).csv")

SouthDakota <- SouthDakota |>
  select(Year, Value)

Alabama <- read_csv("file (30).csv")

Alabama <- Alabama |>
  select(Year, Value)

Alaska <- read_csv("file (31).csv")

Alaska <- Alaska |>
  select(Year, Value)

Arizona <- read_csv("file (32).csv")

Arizona <- Arizona |>
  select(Year, Value)

Arkansas <- read_csv("file (33).csv")

Arkansas <- Arkansas |>
  select(Year, Value)

NewYork <- read_csv("file (34).csv")

NewYork <- NewYork |>
  select(Year, Value)

NorthCarolina <- read_csv("file (35).csv")

NorthCarolina <- NorthCarolina |>
  select(Year, Value)

Oregon <- read_csv("file (36).csv")

Oregon <- Oregon |>
  select(Year, Value)

Pennsylvania <- read_csv("file (37).csv") 

Pennsylvania <- Pennsylvania |>
  select(Year, Value)

California <- read_csv("file (38).csv")

California <- California |>
  select(Year, Value)

Colorado <- read_csv("file (39).csv")

Colorado <- Colorado |>
  select(Year, Value)

Tennessee <- read_csv("file (40).csv") 

Tennessee <- Tennessee |>
  select(Year, Value)

Texas <- read_csv("file (41).csv")

Texas <- Texas |>
  select(Year, Value)

Connecticut <- read_csv("file (42).csv")

Connecticut <- Connecticut |>
  select(Year, Value)

Delaware <- read_csv("file (43).csv")

Delaware <- Delaware |>
  select(Year, Value)

Florida <- read_csv("file (44).csv")

Florida <- Florida |>
  select(Year, Value)

Georgia <- read_csv("file (45).csv")

Georgia <- Georgia |>
  select(Year, Value)

Hawaii <- read_csv("file (46).csv")

Hawaii <- Hawaii |>
  select(Year, Value)

Idaho <- read_csv("file (47).csv")

Idaho <- Idaho |>
  select(Year, Value)

Illinois <- read_csv("file (48).csv")

Illinois <- Illinois |>
  select(Year, Value)

Nebraska <- read_csv("file (49).csv")

Nebraska <- Nebraska |>
  select(Year, Value)

Nevada <- read_csv("file (50).csv")

Nevada <- Nevada |>
  select(Year, Value)

US_data <- bind_rows(list(Alabama = Alabama, Alaska = Alaska, Arizona = Arizona, 
                          Arkansas = Arkansas, California = California, 
                          Colorado = Colorado, Connecticut = Connecticut, 
                          Delaware = Delaware, Florida = Florida, 
                          Georgia = Georgia, Hawaii = Hawaii, Idaho = Idaho, 
                          Illinois = Illinois, Indiana = Indiana, Iowa = Iowa, 
                          Kansas = Kansas, Kentucky = Kentucky, Louisiana = Louisiana, 
                          Maine = Maine, Maryland = Maryland, Massachusetts = Massachusetts, 
                          Michigan = Michigan, Minnesota = Minnesota, Mississippi = Mississippi, 
                          Missouri = Missouri, Montana = Montana, Nebraska = Nebraska, Nevada = Nevada, 
                          "New Hampshire" = NewHampshire, "New Jersey" = NewJersey, "New Mexico" = NewMexico, 
                          "New York" = NewYork, "North Carolina" = NorthCarolina, "North Dakota" = NorthDakota, 
                          Ohio = Ohio, Oklahoma = Oklahoma, Oregon = Oregon, Pennsylvania = Pennsylvania, 
                          "Rhode Island" = RhodeIsland, "South Carolina" = SouthCarolina, "South Dakota" = SouthDakota, 
                          Tennessee = Tennessee, Texas = Texas, Utah = Utah, Vermont = Vermont, 
                          Virginia = Virginia, Washington = Washington, "West Virginia" = WestVirginia, Wisconsin = Wisconsin, 
                          Wyoming = Wyoming), 
                     .id = "name")

head(US_data, 5)


## Data wrangling

state_polygons <- ne_download(type = "states") |>
  filter(admin == "United States of America")

class(state_polygons)
head(state_polygons, 5)

joined_states <- left_join(state_polygons, US_data, by = "name") |>
  select(name, 
         year = Year, 
         value = Value, 
         geometry)
head(joined_states, 5)

## 2012 Data

joined_states_2012 <- joined_states |>
  filter(year == 2012)

categories <- c(0, 60, 65, 70, 75, 80, 85, 90, 95, 100)

colors_2012 <- colorBin("GnBu", domain = joined_states_2012$value, bins = categories)

info_2012 <- sprintf(
  "<strong>%s</strong><br/>%g",
  joined_states_2012$name, joined_states_2012$value) |> 
  lapply(htmltools::HTML)

## 2025 Data

joined_states_2025 <- joined_states |>
  filter(year == 2025)
head(joined_states_2025)


categories <- c(0, 60, 65, 70, 75, 80, 85, 90, 95, 100)

colors_2025 <- colorBin("GnBu", domain = joined_states_2025$value, bins = categories)

info_2025 <- sprintf(
  "<strong>%s</strong><br/>%g",
  joined_states_2025$name, joined_states_2025$value) |> 
  lapply(htmltools::HTML)

## Entire US Data

entire_US <- read.csv("entire_us_data.csv") |>
  drop_na() |>
  group_by(Year) |>
  summarize(
    value = mean(as.numeric(Value))
  )

```

\[Need to do: Introduce what the gender pay gap is and why it matters. I will use other sources I have found (with citations) in this section.\]

This article attempts to visualize the gender pay gap in the United States by using data from the U.S. Bureau of Labor Statistics' [Current Population Survey](https://www.bls.gov/cps/). This survey collects data on full-time U.S. workers' weekly earnings, including data on the median earnings of men and women. These visualizations use fifty-one data sets from this survey (collected from all fifty states) entitled "Women's earnings as percent of men's, Median weekly earnings of full-time and salary workers."[^1]

[^1]: All data sets were downloaded from the U.S. Bureau of Labor Statistics' Data Finder. As of May 27, 2026, all fifty-one data sets (one per U.S. state and one of the entire country) were available on their website.

## Overall U.S. Trends

As we start digging into this data, one introductory question to ask is: ***how has the pay gap evolved over the past few decades?*** Unfortunately, the Census Bureau data is not available before 1979, but, nonetheless, data from the years 1979 to 2026 show some interesting trends.

```{r}
#| label: U.S. Data Linegraph
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| warning: false

ggplot(data = entire_US, 
       mapping = aes(x = Year, y = value)) +
  geom_line(color = "lightblue3", linewidth = 1) +
  labs(
    y = "Percent (%)",
    x = "",
    title = "U.S. data from 1979-2026") +
  theme_minimal() +
  theme(
    axis.title.y = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    plot.title = element_text(size = 18)) +
  transition_reveal(Year, range = NULL, keep_last = TRUE)

```

As a result of some natural variation in the data, there are some small back-and-forth movement over the years. Overall, however, the percent climbed fairly rapidly from approximately 1980 to 2010, stabilizing over 2010 to 2026 at somewhere near 85%. Looking at this visualization, it doesn't appear that the percentage has changed much in the past few years.

However, this doesn't tell us much about what the data looks like at a state level. For example, ***what is the range of percentages across the country? Do they differ geographically by state or region? Or are all parts of the country basically equal in this respect?***

Digging deeper into this observation, the data was broken down into states for both 2012 and 2025.

### 2012 Data (left) vs 2025 Data (right)

```{r}
#| label: percent-graphs
#| echo: false
#| layout-ncol: 2
#| column: screen-inset-shaded


leaflet(joined_states_2012) |>
  addTiles("MapBox") |>
  addPolygons(fillColor = ~colors_2012(value),
                 weight = 1, 
                 opacity = 1,
                 color = "white",
                 dashArray = "4",
                 fillOpacity = 0.8,
              highlightOption = highlightOptions(
                weight = 3, 
                color = "grey",
                dashArray = "4", 
                fillOpacity = 1, 
                bringToFront = TRUE),
              label = info_2012,
              labelOptions = labelOptions(
                style = list("font-weight" = 
                               "normal", 
                             padding = "3px 5px"),
                textsize = "14px",
                direction = "auto")) |>
  addLegend(pal = colors_2012, values = ~value, opacity = 0.8, title = "Percent", position = "bottomright")

leaflet(joined_states_2025) |>
  addTiles("MapBox") |>
  addPolygons(fillColor = ~colors_2025(value),
              weight = 1, 
              opacity = 1,
              color = "white",
              dashArray = "4",
              fillOpacity = 0.8,
              highlightOption = highlightOptions(
                weight = 3, 
                color = "grey",
                dashArray = "4", 
                fillOpacity = 1, 
                bringToFront = TRUE),
              label = info_2025,
              labelOptions = labelOptions(
                style = list("font-weight" = 
                               "normal", 
                             padding = "3px 5px"),
                textsize = "14px",
                direction = "auto")) |>
  addLegend(pal = colors_2025, values = ~value, opacity = 0.8, title = "Percent", position = "bottomright")
```

\[Need to do: (1) explain/unpack the two visualizations above, (2) discuss any lingering concerns/comments about the data\]

